Building and Checking a Suffix Array by Induced Sorting

### 2017-08-26 new-draft/draft.tex
2017/08/27

A suffix array can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods for checking a suffix array when it is being built by any IS suffix sorter, i.e. build and check a suffix array by the IS method simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. By combining the Karp-Rabin fingerprinting technique with our methods, we design two algorithms that perform checking correctly with a negligible error probability. The algorithm designed by the first method has an integer sorting complexity, while the algorithm designed by the second method runs in linear time. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS and implement it for performance analysis. From our experiments, the checking overhead is considerably smaller than the building consumption.

%A suffix array can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods to enable the IS to check a suffix array when it is built, i.e. build and check a suffix array simultaneously. In particular, the first method can be exteneded to check both suffix and LCP arrays. By combining the Karp-Rabin fingerprinting techinque with our methods, we design two algorithms that can check correctly with a negligible error probability. The algorithm designed by the first method has an integer sorting complexity, while the algorithm designed by the second method runs in linear time. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS and implement it for performance evaluation. From our experiments, the checking overhead is considerably smaller than the building consumption.

@@

{\color{red}

The LCP-value $\ell$ of ${\sf suf}(i)$ and ${\sf suf}(j)$ is right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$. In the accepted paper, we check equality of ${\sf sub}(i, i + \ell - 1)$ and ${\sf sub}(j, j + \ell - 1)$ by comparing their fingerprints. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is determined by comparing $x[i + \ell]$ and $x[j + \ell]$.

In this paper, both the lexical order and the LCP-value of two neighboring suffixes in the given suffix array is checked at the top recursion level following the IS principle.

}

@@

A suffix array~(SA) can be built within linear time and space by the internal-memory algorithm SA-IS~\cite{Nong11} using the induced sorting method. According to the IS principle, the order of two suffixes is determined by comparing their heading characters and successors in sequence, where the order of their successors is determined recursively. Recently, the IS method has been also applied to designing three disk-based suffix sorting algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, where eSAIS can build both suffix and LCP arrays at the same time. Compared to the others designed by different methods (e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}), these algorithms runs fast with linear I/O complexity. However, these algorithms currently suffer from a bottleneck due to the large disk space for obtaining the heading characters of unsorted suffixes and the ranks of their sorted successors in a disk-friendly way. It is reported that the average disk use for constructing an SA encoded by 40-bit integers for pSAscan is only $7.5n$ bytes, while that for eSAIS, DSA-IS and SAIS-PQ are $24n$, $18n$ and $15n$ bytes, respectively, where $n$ is the size of input string. The poor space performance observed from these IS algorithms is mainly because the current programs for these algorithms fail to immediately free the disk space for temporary data even when the data is no longer needed. A dramatic improvement can be achieved by storing temporary data in multiple files and deleting a file when the data in the file is not needed any more. This trick has been used by fSAIS~\cite{Karkkainen2017} to engineer the external-memory IS method for better performance, where the peak disk use is $8n$ bytes for constructing an SA of 40-bit integers in the given experiments.

While the research on developing various SA builders using the IS method is evolving, the developed software are becoming complex and appear as open-source without guarantee. A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. The software packages for DC3 and eSAIS provide a checker based on the idea presented in~\cite{Dementiev2008a}. When running on external-memory model, the cost taken  by this checker is rather high as it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. In this paper, we propose two methods to enable the IS method to build and check an SA simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. We employ the Karp-Rabin fingerprinting technique~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. For analysis, we first use new substring sorting and naming methods to improve the design of DSA-IS and then combine the second checking method into the adapted DSA-IS for evaluating the checking overhead. Our experimental results indicate that the time, space and I/O volume for checking SA is considerably smaller than that for building SA.

@@

Characters in $x$ are classified into two categories: L-type and S-type. We say $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] = x[i + 1]$ and $x[i + 1]$ is S-type; otherwise $x[i]$ is L-type. Further, if $x[i]$ and $x[i - 1]$ are respectively S-type/L-type and L-type/S-type, then $x[i]$ is also called S*-type/L*-type. We use an array $t$ to record the type of all the characters in $x$, where $t[i] = 1$ or $0$ if $x[i]$ is S-type or L-type, respectively. The type of a substring or suffix is determined by that of its heading character.

@@

Partition $x$ into multiple S*-type substrings, each substring contains two S*-type characters and any two neighboring substrings overlap an S*-type character. We produce a reduced string $x_1$ by replacing each substring with its name, where the name represents the rank of this	 substring among all. In this paper, we use "rank" and "name" interchangeably to indicate the lexical order of substrings.

@@

All the suffixes in $sa$ are naturally grouped into multiple buckets. Each bucket occupies a contiguous interval in $sa$ and contains all the suffixes with a same heading character. Without loss of generality, we denote by ${\sf sa\_bkt}(c_0)$ the bucket for suffixes starting with $c_0$, where ${\sf sa\_bkt}(c_0)$ can be naturally divided into two parts ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$ containing L-type and S-type suffixes, respectively. We also define ${\sf lcp\_bkt}(c_0)$, ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ on $lcp$ for $c_0 \in \Sigma$.

@@

	/* Check Recursion Condition */ \\	
	\If{exist two equal characters in $x_1$}{

@@

During the execution of a reduction/induction phase, all the substrings/suffixes are sorted by comparing their heading characters and the ranks of their successors according to the IS principle. This involves a great number of random accesses to $x$ and $sa$, which can be done very fast if both $x$ and $sa$ can wholly reside on RAM. However, if the size of input and output exceeds the capacity of internal memory, each access will take an individual I/O operation to disk, leading to a performance degradation. The DSA-IS algorithm solves this problem by performing the following two steps during a reduction/induction phase:

@@

	\item Sort the substrings/suffixes in $x$ by their heading characters and the ranks of their successors in an external-memory heap. \label{dsais_sorting_method:2}
\end{enumerate}

Because all the substrings/suffixes in the heap are sorted already, S\ref{dsais_sorting_method:2} determines the rank of a substring/suffix when popping it from the heap. When inducing the predecessor of current top substring/suffix, the heading character in need is retrieved from external memory by a sequential I/O operation. As shown in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but the former runs slower than the latter due to the large I/O volume for sorting and naming S*-type substrings during the reduction phase. To improve the performance, we propose new substring sorting and naming methods in the following.

@@

All the S*-type substrings are classified into long and short categories with respect to whether or not containing more than $D$ characters. The new substring sorting method mainly consists of the three steps below:

\begin{enumerate}[S1']
	\item Reuse S\ref{dsais_sorting_method:1} to sort the short and the long in each block. Copy the heading characters for sorting the long to external memory in their access order and copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Reuse S\ref{dsais_sorting_method:2} to sort the long in $x$. Copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

	\item Merge the short and the long by a multi-way sorter.~\label{dsaism_sorting_method:3}
\end{enumerate}

The short S*-type substrings in each block are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:1}', while the long S*-type substrings in $x$ are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:2}'. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to determine the lexical order of substrings belonging to different sequences. The heap contains at most $k + 1$ substrings at any point of time. It performs a literal string comparison to compare any two substrings in $\mathcal{O}(D)$\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}.

The above sorting method has a good performance if $D$ is small and the majority of S*-type substrings in $x$ are short. This is commonly satisfied in real-world datasets. We next describe a method for naming the S*-type substrings when they are being sorted according to S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:3}'. The key point is to check equality of two substrings successively popped from the heap. For this, we literally compare the two substrings in $\mathcal{O}(D)$ time if either of them is short; otherwise, we compare their names in $\mathcal{O}(1)$ time instead. Recall that the rank of each long S*-type substring is determined when inducing its predecessor, this value is reused to represent its name herein\footnote{A similar technique has been used to merge the substring sorting and naming processes in SAIS-PQ.}. Our experiments in Section~\ref{sec:experiments} show that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

\begin{IEEEproof} If condition (1) is true, then $sa$ is a permutation of all the suffixes in $x$. If condition (2) is true, then the lexical order of any two neighboring suffixes in $sa$ is determined by comparing their heading characters and the ranks of their successors. As a result, all the suffixes in $x$ are sorted according to the lexical order of their heading characters and successors.

@@

Lemma~\ref{lemma:2} shows that we can check the correctness of $sa^*$ instead of $sa$ at the top recursion level to perform SA verification if the first condition of Lemma~\ref{lemma:2} is always true. In fact, the code snippet for the induction phase at the top recursion level only consists of tens of C++ code lines. This is considerably smaller than the whole program for a disk-based suffix sorting algorithm. Following the idea, we assume S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented and propose two methods for checking computation errors caused by implementation bugs and other malfunctions. The algorithms designed by these methods can be seamlessly integrated into any IS suffix sorting algorithm for building and checking an SA at the same time.

@@

Method A is based on Lemma~\ref{lemma:3}, which is an extension of Lemma~\ref{lemma:2}.

\begin{lemma} \label{lemma:3}
	Assume the induction phase at the top recursion level is correctly implemented, the output $sa[0, n)$ of an IS suffix sorter is the SA for the input string $x[0,  n)$ if the following conditions are satisfied: \\
	(1) $sa^*$ is correct. \\
	(2) $sa[i]$ is equal to the value calculated by S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' for $i \in [0, n)$. \\

\end{lemma}

The first condition of Lemma~\ref{lemma:3} can be checked by a sparse SA checker like~\cite{wu2017}. The second condition intends for detecting malfunctions other than implementation bugs (e.g., I/O errors). This is checked by determining whether the induced and scanned values for each suffix are equal\footnote{Notice that each suffix induced into $sa$ will be latter scanned for inducing the order of its predecessor during S\ref{induction_phase:3}'' and S\ref{induction_phase:4}''. }. The problem to be solved here is that when a suffix is induced into a bucket, its corresponding value in $sa$ may not be scanned at once. Our solution is to ensure the sequence of values placed at a bucket is identical to that scanned latter. For the purpose, we use a fingerprinting function to increasingly compute the fingerprints of both sequences and check their equality in constant time at the end of the induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ can be built and probabilistically checked at the same time.

It was reported in~\cite{Fischer11} and~\cite{Bingmann12} that the IS method can be used to construct LCP array as well. For any two neighboring suffixes in $sa$, their LCP-value is zero if they start with different characters; otherwise, the LCP-value of them is one greater than that of their successors. This implies the possibility for extending Method A to check both suffix and LCP arrays. We design Algorithm~\ref{alg:2} to check $sa$ and $lcp$ when they are induced from $sa^*$ and $lcp^*$ at the top recursion level of an IS builder. In this algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$. These values are compared for checking the second condition of Lemma~\ref{lemma:3}. The algorithm also computes and compares ${\sf lcp_{I1}}(c)/{\sf lcp_{I2}}(c)$ and ${\sf lcp_{S1}}(c)/{\sf lcp_{S2}}(c)$ in the meantime.

@@

According to Lemma~\ref{lemma:4}, Method B uses a different way to check the correctness of $sa^*$. Before our presentation, we first introduce a notation called $\overline{sa^*}$. The same as $sa^*$, $\overline{sa^*}$ also represents the suffix array for all the S*-type suffixes. Both $sa^*$ and $\overline{sa^*}$ are computed during the induction phase at the top recursion level, where the former is calculated from $sa_1$ in S\ref{induction_phase:1}'' and the latter is calculated when inducing the S*-type suffixes into $sa$ in S\ref{induction_phase:4}''.

@@

	(1) $sa^*[i]$ is a permutation of all the S*-type suffixes. \\
	(2) $sa^*[i] = \overline{sa^*}[i]$ for $i \in [0, n_1)$. \\
	(3) $sa[i]$ is equal to the value calculated by S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' for $i \in [0, n)$. \\

\end{lemma}

@@

The first condition of Lemma~\ref{lemma:4} is naturally satisfied when computing $sa^*$ from $sa_1$ in S\ref{induction_phase:1}''. The second condition is probabilistically checked by computing and comparing the fingerprints of $sa^*$ and $\overline{sa^*}$. The fingerprint of $sa^*$ is computed when placing the elements of $sa^*$ into $sa$ in S\ref{induction_phase:2}'', while the fingerprint of $\overline{sa^*}$ is computed when inducing the S*-type suffixes into $sa$ in S\ref{induction_phase:4}''. We design Algorithm~\ref{alg:3} to check if the final SA observes Lemma~\ref{lemma:4}.

@@

        Compute the fingerprints of $sa^*$, ${\sf sa_{I1}}(c)$ and ${\sf sa_{S1}}(c)$ when inducing the order of L-type suffixes. \\

        Check if ${\sf sa_{I1}}(c) = {\sf sa_{S1}}(c)$ by comparing their fingerprints. \\

        Compute the fingerprints of $\overline{sa^*}$, ${\sf sa_{I2}}(c)$ and ${\sf sa_{S2}}(c)$ when inducing the order of S-type suffixes. \\

        Check if ${\sf sa_{I2}}(c) = {\sf sa_{S2}}(c)$ by comparing their fingerprints. \\

@@

The proposed two methods check equality of two arrays (or sub-arrays) by comparing their fingerprints, which can be calculated using a hash function. Notice that two equal arrays must be mapped to an identical hash value, but the inverse is not always true. To lower the error probability of a false match, we prefer using the Karp-Rabin fingerprinting function to compute these hash values, called fingerprints,	 following Formulas~\ref{formula:1}-\ref{formula:2}, where $L$ is a prime and $\delta$ is an integer randomly chosen from $[1, L)$. By setting $L$ to a large value, the error probability can be reduced to a negligible level. In Fig.~\ref{fig:example2}, we depict an example for computing the integer array $A$ identical to $sa^*$ and $\overline{sa^*}$ in Fig.~\ref{fig:example1}. Given $L = 23$ and $\delta = 5$, we first compute the fingerprint of $A[0, 0]$ in line 4. Because $\sf fp(A[0, -1]) = 0$, $\sf fp(A[0, 0]) = (0 \cdot 5 + 14) \, mod \, 23 = 14$. By iteratively computing the fingerprints of the prefixes of $A$, we finally obtain the fingerprint of $A$ in line 7.

@@

A key operation of Algorithms~\ref{alg:2} and~\ref{alg:3} is to check equality of two integer sequences by comparing their fingerprints. This is done in linear time and constant RAM space by using the Karp-Rabin fingerprinting technique. Algorithm~\ref{alg:2} also employs the method presented in~\cite{wu2017} to verify $sa^*$ and $lcp^*$. For any two neighboring suffixes in $sa^*$, this method checks their lexical order indicated by $sa^*$ and the LCP-value $\ell$ indicated by $lcp^*$ as follows: (1) check if the fingerprints of their leftmost $\ell$ characters are equal; (2) check if their $(\ell + 1)$-th characters are different; (3) check if their lexical order corresponds to that of their $(\ell + 1)$-th characters. It should be noticed that this method checks all the neighboring suffix pairs in batch within a sorting complexity. Because at most one out of every two suffixes is S*-type and the checking process is only executed at the top recursion level, the consumption for checking $sa^*$ and $lcp^*$ is much less than that for building the two arrays by the IS method.

@@

By assuming the induction phase at the top recursion level is correctly implemented, we proposed two methods that enable any IS suffix sorter to build and check an SA simultaneously. The probabilistic algorithm designed by the second method is rather lightweight, it takes negligible time and space compared with the existing IS suffix sorting and checking algorithms. In this paper, we also made the first attempt to improve the performance of DSA-IS using new substring and naming methods. Our program for the adapted algorithm DSA-IS+ runs as fast as that for eSAIS and consumes only half as much disk space as the latter on various real-world datasets. We are now designing and implementing a novel IS suffix sorter that takes no more than 1n work space on external memory model. Theoretically, this suffix sorter has a better space performance than fSAIS under the same circumstances.

>>>>

A suffix array~(SA) can be built on both internal and external memory models following the induced sorting~(IS) principle. We propose two methods for checking an SA when it is being built by any IS suffix sorter. The first method can be used to probabilistically check both suffix and longest common prefix~(LCP) arrays, while the second method is for checking SA deterministically. To achieve a good performance, we combine the Karp-Rarbin fingerprinting techinque into our methods to design two algorithms that perform checking correctly with a negligible error probability. The algorithm designed by the first method has an integer sorting complexity, while the algorithm designed by the second method runs within linear time and constant RAM space. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS for evaluation, where the experimental results indicate that the checking overhead is considerably smaller than the building consumption.

@@

@@

The SA-IS algorithm takes linear time and RAM space to build an SA on internal memory model~\cite{Nong11}. The IS method determines the lexical order of two suffixes by comparing their heading characters and successors in sequence, where the lexical order of their successors is determined recursively following the same way. 
Recently, the IS method has been applied to designing three disk-based suffix sorting algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}. In particular, eSAIS can build both suffix and LCP arrays at the same time. 
These IS algorithms have linear I/O complexity better than the others designed by different methods (e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}), but they suffer from a bottleneck due to the large disk space for obtaining the heading characters of unsorted suffixes and the ranks of their sorted successors in a disk-friendly way. 
As reported, the average disk use for pSAscan to build a size-$n$ SA encoded by 40-bit integers is around $7.5n$ bytes, while that for eSAIS, DSA-IS and SAIS-PQ are $24n$, $18n$ and $15n$ bytes, respectively. 
The poor space performance for these IS algorithms is mainly because their current programs fail to free the disk space for temporary data even when the data is no longer needed. 
A dramatic improvement can be achieved by storing temporary data into multiple files and deleting a file immediately when the data in it is not needed anymore. This trick has been employed by fSAIS~\cite{Karkkainen2017} to engineer the external-memory IS method, where the peak disk use is $8n$ bytes for building an SA of 40-bit integers in the given experiments.

While the research on developing various SA builders using the IS method is evolving, the developed software are becoming complex and appear as open-source without guarantee. 
A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. 
The software packages for DC3 and eSAIS provide a checker based on the idea presented in~\cite{Dementiev2008a}. 
When running on external-memory model, the cost taken by this checker is rather high as it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. 
We propose in this paper two methods that enable the IS method to build and check an SA at the same time. 
Our first method is probabilistic and can be extended to check both suffix and LCP arrays, while the second method is deterministic and designed for checking SA only. 
By combining the Karp-Rabin fingerprinting technique~\cite{Karp1987}, we design two probabilistic algorithms to perform checking correctly with a negligible error probability. The key operation of both algorithms is to check equality of two integer sequences by computing and comparing their fingerprints. As will be shown, the algorithm designed by the first method checks an SA within a integer sorting complexity and the algorithm designed by the second method checks an SA in linear time. For a further study, we first use new substring sorting and naming methods to improve the design of DSA-IS and then integrate the algorithm designed by the second method into the adapted DSA-IS for evaluation. Our experimental results indicate that the time, space and I/O volume for checking SA is negligible in comparison with that for building SA.

@@

Characters in $x$ are classified into two categories: L-type and S-type. We say $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] = x[i + 1]$ and $x[i + 1]$ is S-type; otherwise $x[i]$ is L-type. Further, if $x[i]$ and $x[i - 1]$ are respectively S-type/L-type and L-type/S-type, then $x[i]$ is also called S*-type/L*-type. We use an array $t$ to record the type of all the characters in $x$, where $t[i] = 1$ or $0$ if $x[i]$ is S-type or L-type, respectively. The type of a substring or suffix is determined by that of its heading character. In the rest of this paper, we only consider substrings ending with an S*-type character.

@@

Partition $x$ into multiple S*-type substrings, each substring contains two S*-type characters and any two neighboring substrings overlap an S*-type character. We produce a reduced string $x_1$ by replacing each substring with its name, where the name represents the rank of this	 substring among all. We interchangeably use "rank" and "name"  to indicate the lexical order of a substring.

@@

All the suffixes in $sa$ are naturally grouped into multiple buckets. Each bucket occupies a contiguous interval in $sa$ and contains all the suffixes with a same heading character. Without loss of generality, we denote by ${\sf sa\_bkt}(c_0)$ the bucket for suffixes starting with $c_0$, where ${\sf sa\_bkt}(c_0)$ can be naturally divided into two parts ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$ that contain L-type and S-type suffixes, respectively. Similarly, we define ${\sf lcp\_bkt}(c_0)$, ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ on $lcp$ for $c_0 \in \Sigma$.

@@

	/* Check Recursion Condition */ \\	
	\If{exist duplicate characters in $x_1$}{

@@

During the execution of a reduction/induction phase, all the substrings/suffixes are sorted by comparing their heading characters and the ranks of their successors according to the IS principle. This involves a great number of random accesses to $x$ and $sa$, which can be done very fast if both $x$ and $sa$ can wholly reside on RAM. However, if the size of input and output exceeds the capacity of internal memory, each access may take an individual I/O operation to disk, leading to a performance degradation. The DSA-IS algorithm solves this problem by executing the following two steps during a reduction/induction phase:

@@

	\item Sort the substrings/suffixes in $x$ according to the lexical order of their heading characters and the ranks of their successors by using an external-memory heap. \label{dsais_sorting_method:2}
\end{enumerate}

S\ref{dsais_sorting_method:2} induces the lexical order of substrings/suffixes by using an external-memory heap. It continually pops the top substring/suffix from the heap and induces its predecessors into the heap according to the rank of the top substring/suffix and the heading character of the predecessors. For all the induced substrings/suffixes, their heading characters are already arranged in access order on disk by S\ref{dsais_sorting_method:1} and thus can be retrieved from external memory by sequential I/O operations before use. As shown in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but the former runs slower than the latter due to the large I/O volume for sorting and naming S*-type substrings during the reduction phase. To improve the performance, we propose new substring sorting and naming methods in the following.

@@

All the substrings are classified into long and short categories with respect to whether or not containing more than $D$ characters. The new substring sorting method mainly consists of the three steps below:

\begin{enumerate}[S1']
	\item Reuse S\ref{dsais_sorting_method:1} to sort all the substrings in each block. Copy the heading characters of the induced long substrings to external memory in access order and copy the induced short S*-type substrings to external memory in sorted order.~\label{dsaism_sorting_method:1}

	\item Reuse S\ref{dsais_sorting_method:2} to sort the long substrings in $x$. Copy the leftmost $D$ characters of the long S*-type substrings to external memory in sorted order.~\label{dsaism_sorting_method:2}

	\item Merge all the short and long by a multi-way sorter.~\label{dsaism_sorting_method:3}
\end{enumerate}

The short S*-type substrings in each block are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:1}', while the long S*-type substrings in $x$ are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:2}'. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to determine the lexical order of substrings belonging to different sequences. The heap contains at most $k + 1$ substrings at any point of time. It performs a literal string comparison to compare any two S*-type substrings in $\mathcal{O}(D)$\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}.

The above sorting method has a good performance if $D$ is small and the majority of S*-type substrings in $x$ are short. This is commonly satisfied in real-world datasets. We next describe a method for naming the S*-type substrings when they are being sorted by S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:3}'. The key operation is to check equality of two substrings successively popped from the heap in S\ref{dsaism_sorting_method:3}'. If either of the substrings is short, then we literally compare them in $\mathcal{O}(D)$ time. Otherwise, we check whether their ranks are equal by comparing their heading characters and the ranks of their successors in S\ref{dsaism_sorting_method:2}'. For the same purpose, SAIS-PQ also applies a similar technique to merge the substring sorting and naming processes. Our experiments in Section~\ref{sec:experiments} show that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

\begin{IEEEproof} If condition (1) is true, then $sa$ is a permutation of all the suffixes in $x$. If condition (2) is true, then the lexical order of any two neighboring suffixes in $sa$ is determined by comparing their heading characters and the ranks of their successors. As a result, all the suffixes in $x$ are sorted following the induced sorting principle.

@@

Lemma~\ref{lemma:2} shows that we can check the correctness of $sa^*$ instead of $sa$ at the top recursion level to perform SA verification if the first condition of Lemma~\ref{lemma:2} is always true. In fact, the code snippet for the induction phase at the top recursion level only consists of tens of C++ code lines. This is considerably smaller than the whole program for a disk-based suffix sorting algorithm. Following the idea, we assume S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented and propose two methods for checking computation errors caused by implementation bugs and other malfunctions. These methods adopt different ways to ensure the correctness of $sa^*$. Particularly, our first method can be used to check both suffix and LCP arrays.

@@

The induced sorting method can be applied to building LCP array as well~\cite{Fischer11}. That is, the LCP-value of two substrings can be induced from that of their successors as following: the LCP-value is equal to zero if the two substrings have different heading characters; otherwise, the LCP-value is one greater than that of their successors. Assume the induction phase for inducing $sa$ and $lcp$ from $sa^*$ and $lcp^*$ is correctly implemented at the top recursion level, we show in Lemma~\ref{lemma:3} a set of conditions for checking both $sa$ and $lcp$ when they are built by an IS builder like~\cite{Fischer11, Bingmann12}.

\begin{lemma} \label{lemma:3}
	Assume the induction phase at the top recursion level is correctly implemented, the output $sa[0, n)$ and $lcp[0, n)$ of an IS builder are the suffix and LCP arrays for the input string $x[0, n)$ with a high probability if the following conditions are satisfied: \\
	(1) $sa^*$ and $lcp^*$ are correct according to the checking result of the sparse suffix and LCP arrays probabilistic checker~\cite{wu2017}. \\
	(2) $sa[i]$ and $lcp[i]$ are equal to the values induced into the corresponding positions of $sa$ and $lcp$ during the induction phase, for $i \in [0, n)$. \\
\end{lemma}

The checker in~\cite{wu2017} verifies $sa^*$ and $lcp^*$ within an integer sorting complexity. For each pair of neighboring suffixes in the given $sa$, it first employs the Karp-Rabin fingerprinting technique to compute the fingerprints of their leftmost $\ell$ characters, where $\ell$ is their LCP-value indicated by the given $lcp$. If the two fingerprints are equal, then it continues to check if the lexical order of  their $(\ell + 1)$-th characters corresponds to that of themselves indicated by the given $sa$. The second condition in Lemma~\ref{lemma:3} is intended for detecting malfunctions other than implementation bugs. Because each suffix induced into $sa/lcp$ will be latter scanned for inducing the order of its predecessor, this condition can be checked by determining whether the induced and scanned values for each suffix are equal. The problem here is that when a suffix is induced into a bucket, its corresponding value in $sa$ may not be scanned at once. Our solution is to ensure the sequence of values placed at a bucket is identical to that scanned latter. For the purpose, we reuse the Karp-Rabin fingerprinting technique to increasingly compute the fingerprints of both sequences and check their equality in constant time at the end of induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ and $lcp$ can be built and probabilistically checked at the same time.

We design Algorithm~\ref{alg:2} to check $sa$ and $lcp$ when they are induced from $sa^*$ and $lcp^*$ at the top recursion level of an IS builder. In this algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$. The algorithm computes and compares the fingerprints of these sequences and those of ${\sf lcp_{I1}}(c)/{\sf lcp_{I2}}(c)$ and ${\sf lcp_{S1}}(c)/{\sf lcp_{S2}}(c)$ to check the second condition of Lemma~\ref{lemma:3}.

@@

Method B is deterministic and specific for checking SA only. It uses a different way to ensure the correctness of $sa^*$. Before our presentation, we first introduces a notation called $\overline{sa^*}$. The same as $sa^*$, $\overline{sa^*}$ also represents the suffix array for all the S*-type suffixes. Both $sa^*$ and $\overline{sa^*}$ are computed during the induction phase at the top recursion level, where the former is calculated from $sa_1$ in S\ref{induction_phase:1}'' and the latter is computed when inducing the S*-type suffixes into $sa$ in S\ref{induction_phase:4}''.

@@

	(1) $sa^*[i]$ is a permutation of all the S*-type suffixes. \\
	(2) $sa^*[i] = \overline{sa^*}[i]$ for $i \in [0, n_1)$. \\
\end{lemma}

@@

The first condition of Lemma~\ref{lemma:4} is naturally satisfied when computing $sa^*$ from $sa_1$ in S\ref{induction_phase:1}''. One way for deterministically checking the second condition is to copy both $sa^*$ and $\overline{sa^*}$ to external memory and compare their elements one by one after the induction phase. Algorithm~\ref{alg:3} applies a more space efficient way to checking equality of the two integer sequences with a negligible error probability. It reuses the Karp-Rabin fingerprinting technique to compute and compare the fingerprints of $sa^*$ and $\overline{sa^*}$, where the fingerprint of $sa^*$ is computed when placing the elements of $sa^*$ into $sa$ in S\ref{induction_phase:2}'' and the fingerprint of $\overline{sa^*}$ is computed when inducing the S*-type suffixes into $sa$ in S\ref{induction_phase:4}''.

@@

        Compute the fingerprints of $sa^*$ when inducing the order of L-type suffixes. \\

        Compute the fingerprints of $\overline{sa^*}$ when inducing the order of S-type suffixes. \\

@@

Both Algorithms~\ref{alg:2} and ~\ref{alg:3} check equality of sequences by comparing their fingerprints. We can use any rolling hash function to compute the fingerprints in need. Notice that two equal sequences must be mapped to an identical hash value, but the inverse is not always true. To lower the error probability of a false match, we prefer using the Karp-Rabin fingerprinting techniques to compute these values following Formulas~\ref{formula:1}-\ref{formula:2}, where $L$ is a prime and $\delta$ is an integer randomly chosen from $[1, L)$. By setting $L$ to a large value, the error probability can be reduced to a negligible level. In Fig.~\ref{fig:example2}, we depict an example for computing the integer array $A$ identical to $sa^*$ and $\overline{sa^*}$ in Fig.~\ref{fig:example1}. Given $L = 23$ and $\delta = 5$, we first compute the fingerprint of $A[0, 0]$ in line 4. Because $\sf fp(A[0, -1]) = 0$, $\sf fp(A[0, 0]) = (0 \cdot 5 + 14) \, mod \, 23 = 14$.  By iteratively computing the fingerprints in this way, we finally obtain the fingerprint of $A$ in line 7.

@@

Algorithms~\ref{alg:2} and ~\ref{alg:3} take linear time and constant RAM space for computing the target sequences by the Karp-Rabin fingerprinting technique. Algorithm~\ref{alg:2} also employs the checker presented in~\cite{wu2017} to verify $sa^*$ and $lcp^*$ within an integer sorting complexity. Because at most one out of every two suffixes is S*-type and the checking process is only executed at the top recursion level, the overhead for checking $sa^*$ and $lcp^*$ is considerably smaller than that for constructing the two arrays by an IS builder.

@@

By assuming the induction phase at the top recursion level is correctly implemented, we propose two methods that enable an IS builder to build and check an SA simultaneously. The probabilistic algorithm designed by Method B is rather lightweight, it takes negligible time and space to run compared to the existing IS suffix sorting and checking algorithms. We also made our first attempt to improve the performance of DSA-IS using new substring sorting and naming methods. Our program for the adapted algorithm DSA-IS+ runs as fast as that for eSAIS and consumes only half as much disk space as the latter on various real-world datasets. We are now designing and implementing a novel IS suffix sorter that takes no more than 1n work space on external memory model. Theoretically, this suffix sorter has a better space performance than fSAIS under the same circumstances.
